Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-21 endofquarterpaper.qmd

We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and boundaries? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each phase using seriation and network analyses. We then test the strength of the clusters using modularity and perMANOVA tests. Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each phase.

@@

artifact_count_median <- median(artifact_count_df$artifact_count)
```

@@

@@

The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

@@

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R2 statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, which compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups, with larger pseudo F-statistics indicating more separation between groups, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).

<!-- SC: I think we need to elaborate on what the null hypothesis is here. This blog, https://archetypalecology.wordpress.com/2018/02/21/permutational-multivariate-analysis-of-variance-permanova-in-r-preliminary/, does a pretty good job of explaining what exactly the p-value means, which is the proportion of F-statistics for the permuted tests that are greater than or equal to the actual F-statistic.-->

The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

## Effects of Geographic Distance Using Mantel Tests

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. If cultural difference can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. To test this, we determined the correlation between sign distribution and geography using Mantel tests, run with a geographic distance matrix and a Jaccard similarity matrix. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]

## Cultural Group Detection

@@

```{r fig.height= 9, fig.width= 13}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"

@@

For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types, and one with a broader range. The notch and line signs are the most prevalent sign types across all time periods. In the Transitional phase, there is a small range of sign types. The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader group consists of Labeko Koba, Hohle Fels, and El Castillo. In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have notches and lines, and one also has a dashline. The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase. In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs. The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase. We also see that simpler signs, like line and notch, tend to be more common in both groups, but tend to primarily make up the restricted groups.

<!-- BM: I reckon we need to say something about the signs also, not just the sites -->

@@

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase, as shown in @tbl-group-table. The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.

```{r}
 #| label: tbl-group-table
 #| tbl-cap: "Table of the sites by group per phase"
 #| tbl-cap-location: bottom
 group_table <- read_excel("data/site-groups.xlsx")

 kable(group_table)
```

@@

## Modularity test
trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
early_strength <- strength_function(early_artifact_data, early_unique_data)
evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)

@@

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are fairly minimal. The results tell us that the Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and that the Evolved Aurignacian exhibits the weakest groupings, with a score of `r evolved_modularity`. There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger decrease with the Evolved Aurignacian.

The R2 statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership. The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`. We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods. The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest score of `trans_perm_f`, and the Early Aurignacian having the lowest score of `early_perm_f`. Similar to the R2 statistic, we see a decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period. The perMANOVA p score quantifies the statistical significance of the pseudo-F statistic, with 0.05 as the threshold. The only time period to not meet that threshold is the Transitional period (p = `r trans_perm_p`). There is a slight increase between the Proto and Early Aurignacian, and a slightly large increase in the Evolved Aurignacian period, although they all still meets the threshold for statistical significance.

## Effect of Geographic Distance

```{r}
#| label: tbl-mantel-table
#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
#| tbl-cap-location: bottom
## Mantel test for geography vs signs
mantel_table <- full_table %>% 
  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)

proto_mantel_r <- proto_strength %>% 
  dplyr::select(`Mantel R`)

proto_mantel_p <- proto_strength %>% 
  dplyr::select(`Mantel p`)

early_mantel_r <- early_strength %>% 
  dplyr::select(`Mantel R`)

early_mantel_p <- early_strength %>% 
  dplyr::select(`Mantel p`)

The relevance of the isolation-by-distance hypothesis also varies from period to period. The p score quantifies the correlation, with a score of 0.05 or lower meaning that there is a statistically significant relationship between sign type distribution and geographical distance. None of the periods meet that threshold, meaning that the isolation-by-distance hypothesis is not applicable to any of them. However, we still see changes in the p score between periods, suggesting a higher level of correlation with geographic distance in some, and a lower correlation in others. The Proto-Aurignacian period demonstrates the least amount of correlation (Mantel R = `r proto_mantel_r`, p = `r proto_mantel_p`), while the Early Aurignacian exhibits the highest amount of correlation (Mantel R = `r early_mantel_r`, p = `r early_mantel_p`).

ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf, aes(color = group)) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2,
                  bg.color = "white",
                  bg.r = 0.25) +
  theme_minimal() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank())

@fig-group-map shows the geographic distribution of sites as colored by group, demonstrating that geographic boundaries do not readily correspond to the group distributions. In the Transitional phase, the sites are widely spread out across Western Europe, and tend to be located fairly closely to the other sites in their groups. The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle. In the Proto-Aurignacian, we see a slightly more restricted distribution of sites westward but a spread northward, and more overlapping of sites between groups. For the restricted group, we interestingly see a divide across the center, with Mladeč and Hohlenstein-Stadel located close together further to the east, and Gatzarria and Abri Pataud located close together further to the west, with a large space between the two clusters. The broader sites tend to be more uniformly spread out across the geographic space, without much clustering close together. We again see some overlap between Abri Pataud and La Ferrassie, as well as between Geissenklösterle and Hohlenstein-Stadel. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, incluing a great deal of both intra- and inter-group proximity. In the Evolved Aurignacian, we see a wider distribution and more distance between sites. In particular, the sites in the reduced group are especially spread out across the space, while the broader sites are slightly more clustered. We also see a bit of a geographic divide between groups, with the reduced sites tending to be located further to the west, while the broader sites tend to be located further to the east. The exceptions is Sirgenstein Cave, which is located closer to the broader sites, almost overlapping with Vogelherd, Bockstein-Törle, and Hohle Fels.

@@

@@

>>>>

We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and boundaries? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each phase using seriation and network analyses. We then test the strength of the clusters using modularity and perMANOVA tests. All of our programming in this study was conducted in RStudio.

@@

artifact_count_median <- median(artifact_count_df$artifact_count)
n_material <- c(unique(signbase_full_clean$material))

@@

```{r fig.height= 9, fig.width=10}
```

@@

The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]. The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

@@

@@

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

```{r}
#| label: tbl-mantel-table
#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
#| tbl-cap-location: bottom
## Mantel test for geography vs signs

trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
 proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
 early_strength <- strength_function(early_artifact_data, early_unique_data)
 evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)

mantel_table <- full_table %>% 
  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)

To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value. The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups. Larger pseudo-F statistics indicate more separation between groups. The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

## Cultural Group Detection
```{r fig.height= 9, fig.width= 13}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data, with group membership indicated by color. Restricted group is in yellow, broader group is in blue."

@@

```{r}
sign_type_df <- time_unique_data_full %>% 
  pivot_longer(cols = line:star) %>% 
  group_by(id) %>% 
  filter(value != 0) %>% 
  mutate(value = 1) %>% 
  group_by(id) %>% 
  summarise((sign_count = sum(value))) %>% 
  mutate(sign_type_count = `(sign_count = sum(value))`, `(sign_count = sum(value))` = NULL)

time_unique_data_full <- time_unique_data_full %>% 
  left_join(sign_type_df, join_by(id))

max_signs <- max(time_unique_data_full$sign_type_count)

For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types (1-3 types), which tend to be simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs. Simpler signs composed of fewer components, like line and notch, are the most prevalent across all time periods. Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.

In the Transitional phase, there is a relatively small range of sign types compared to other phases. The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch. The broader group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.

In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases. The restricted group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline. The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase. In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader group.

In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups. The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line). The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations. In addition to line and notch, the obline, dot and vulva are also frequently found in the broader group.

Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase. Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs. The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels. Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader group.

@@

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).

@@

## Modularity test

@@

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal. The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`. There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.

The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership. The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`. We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods. The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`. Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period. The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto- , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).

@@

ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf, aes(color = group)) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2,
                  bg.color = "white",
                  bg.r = 0.25) +
  theme_minimal() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        strip.text = element_text(size = 10))

The geographic structure of the groups changes from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups. The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together. In the Evolved Aurignacian, we see a wider distribution and more distance between sites.

@@

It is also possible that the established precedent of aggregation sites could explain some of our results. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites are places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Soffer_Adovasio_2010; @Bourdier_2013; @Svoboda_2013]. It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose. However, without examining environmental data, this remains only a hypothesis at this time.

Our work adds to a growing awareness of diversity within the Upper Paleolithic. Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @3567b0e7-61f1-3f3c-8f91-c78399bc0a5a]. Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @fbdeb6f7-9945-3e64-82d6-7c3ab51f8115; @97387c23-a43b-34e6-95ef-0a248850bcf3; @c35eb420-0df4-3425-adac-d15b922564bd]. This study contributes to that awareness, suggesting permeable, yet distinct cultural patterning and boundaries, as well as patterns of exchange and movement between groups.

This study also adds to the growing body of methods for the identification of cultural groups, as well as change in social dynamics over time. We have shown that seriation and network analysis can be successfully used to identify cultural groups from geometric signs on mobile objects, and modularity and perMANOVA tests can be used to evaluate the strength of those groupings.

One limitation of the study is the applicability of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups. It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange. Ultimately, we chose to evaluate them as evidence of cultural connection, however, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.

Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity. We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum. This is in accordance with the cutural phylogeny models proposed by @boyd2013cultural, in which cultural groups are defined by presence or absence of certain cultural variants. In this case, restricted sign variability vs. broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.
